An Analysis of Homelessness in Toronto in regards to COVID-19 and other factors

### 2022-02-08 outputs/paper/paper.Rmd
Finish polishing of paper

---
title: "An Analysis of Homelessness in Toronto in regards to COVID-19 and other factors"
subtitle: "Statisical analysis from January 2020 to December 2021"
author: "Ziyao Han"
thanks: "Code and data are available at: https://github.com/HanFrank/Homelessness-Analysis-of-Toronto"
date: "February 6, 2022"
abstract: "A well functioning shelter system and its accompanying services should be an essential part of the nervous system of a well-run city. This report uses data collected from Open Data Toronto and applies statistical visualization, past and present literature to analyse the homelessness situation in Toronto. The data shows that COVID-19 has had a significant impact on shelter usage and is a predominant factor in shelter usage and the homelessness situation in Toronto within the last two years. This report uses a range of data to analyse the shelter system to provide insight into the ongoing fight against homelessness."
output: 
  bookdown::pdf_document2
toc: TRUE
bibliography: references.bib
---

@@

There is a variety of reasons a person experiencing homelessness will not seek help from a shelter system. An example of this could be to avoid shelter restrictions such as curfews. Another potential reason, especially during the Covid-19 pandemic, is public health concerns [@citepallet]. This specific topic will be discussed further in section \@ref(covid).

@@

```{r extract, echo = FALSE, warning = FALSE, message = FALSE, fig.align="center"}
# Showing an extract of the total dataset as a sample
monthly_shelter_usage %>%
  select(date, population_group, actively_homeless, population_group_percentage) %>%
  slice(1:7) %>%
  kable(
    caption = "Extract from full dataset, January 2020",
    col.names = c("Date", "Population Group", "Actively Homeless", "Percentage of Group Population"),
    digits = 1,
    booktabs = TRUE,
    linesep=""
  )
```

@@

```{r graph1, fig.cap="Active number of homeless over time", echo=FALSE, fig.height=4, fig.width=6}
# Creating subset of only total population
total_population <- monthly_shelter_usage[monthly_shelter_usage$population_group == "All Population",]
# Plotting total usage of homeless shelter over time and overlaying start of covid
ggplot(total_population, aes(x = date, y = actively_homeless)) +
geom_line() +
# Drawing dotted red line for covid
geom_line(aes(x=as.Date("2020-03-01"), col = "COVID-19 response begins in Toronto"), linetype="dotted") + 
labs(x = "Date", 
     y = "Actively Homeless", 
     title =  "Usage of Homeless shelters dropping due to COVID-19",
     col = "Event: ") +
theme_minimal() +
theme(legend.position = "bottom")
```

@@

Figure \@ref(fig:graph2) shows the active number of homeless people in each population group of the dataset. Almost all population groups share a similar behaviour, with one notable exception, the chronic homeless population group.

```{r graph2, fig.cap="Active number of homeless over time for each population group",fig.height=4, echo=FALSE}
ggplot(monthly_shelter_usage, mapping = aes(x = date, y = actively_homeless, colour = population_group)) +
geom_line() +
labs(x = "Date", 
     y = "Actively Homeless", 
     title =  "Chronic shelter usage remains stable through COVID-19",
     caption = "Began tracking Indigenous numbers in Jan 2021",) +
theme_minimal()
```

@@

Low female representation in homelessness numbers could be due to a variety of factors. Women and children are often especially concerned about conflict and abuse due to the public nature of homeless shelters. This would be of even greater concern for victims of sexual abuse that are seeking refuge or are temporarily homeless, may intentionally avoid homelessness shelters or services [@citepallet]. This is an extremely expansive topic that could be an area of further exploration as discussed in section \@ref(further).

@@

Figure \@ref(fig:graph3) shows the number of people moving to and from permanent housing during the two year period. The red line indicates the number people leaving the shelter system by obtaining permanent housing and the blue line represents the number of people reentering the shelter system after moving to permanent housing.

```{r graph3, fig.cap="Movement between Permanent housing", echo=FALSE, fig.height=4, fig.width=6}
ggplot(total_population, aes(x = date)) +
geom_line(aes(y = moved_to_housing, col = "Moved")) +
geom_line(aes(y = returned_from_housing, col = "Returned")) + 
# Drawing dotted red line for covid
labs(x = "Date", 
     y = "# of people", 
     title =  "People moving to permanent housing dropping significantly",
     col = "Action: ") +
theme_minimal() +
theme(legend.position = "bottom")
```

>>>>

---
title: "An Analysis of Homelessness in Toronto in regards to COVID-19 and other factors"
subtitle: "Statistical analysis from January 2020 to December 2021"
author: "Ziyao Han"
thanks: "Code and data are available at: https://github.com/HanFrank/Homelessness-Analysis-of-Toronto"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "A well functioning shelter system and its accompanying services should be an essential part of the nervous system of a well-run city. This report uses data collected from Open Data Toronto and applies statistical visualization, past and present literature to analyse the homelessness situation in Toronto. The data shows that COVID-19 has had a significant impact on shelter usage and is a predominant factor in shelter usage and the homelessness situation in Toronto within the last two years. This report uses a range of data to analyse the shelter system to provide insight into the ongoing fight against homelessness."
output: 
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

@@

There is a variety of reasons a person experiencing homelessness will not seek help from a shelter system. An example of this could be to avoid shelter restrictions such as curfews. Another potential reason, especially during the COVID-19 pandemic, is public health concerns [@citepallet]. This specific topic will be discussed further in Section \@ref(covid).

@@

```{r extract, echo = FALSE, warning = FALSE, message = FALSE}
# Showing an extract of the total dataset as a sample
monthly_shelter_usage %>%
  select(date, population_group, actively_homeless, population_group_percentage) %>%
  slice(1:7) %>%
  kable(
    caption = "Extract from full dataset, January 2020",
    col.names = c("Date", "Population Group", "Actively Homeless", "Percentage of Group Population"),
    digits = 1,
    booktabs = TRUE,
    linesep=""
  )
```

@@

```{r graph1, fig.cap="Active shelter usage over time", echo=FALSE, fig.height=5, fig.width=6}
# Creating subset of only total population
total_population <- monthly_shelter_usage[monthly_shelter_usage$population_group == "All Population",]
# Plotting total usage of homeless shelter over time and overlaying start of covid
ggplot(total_population, aes(x = date, y = actively_homeless)) +
geom_line() +
# Drawing dotted red line for covid
geom_line(aes(x=as.Date("2020-03-01"), col = "COVID-19 response begins in Toronto"), linetype="dotted") + 
labs(x = "Date", 
     y = "Actively Shelter Usage", 
     title =  "Usage of Homeless shelters dropping due to COVID-19",
     col = "Event: ") +
theme_minimal() +
theme(legend.position = "bottom") # Putting legend on the bottom for aesthetic reasons 
```

@@

Figure \@ref(fig:graph2) shows the active number of shelter usages in each population group of the dataset. Almost all population groups share a similar behaviour, with one notable exception, the chronic homeless population group.

```{r graph2, fig.cap="Active shelter usage over time for each population group",fig.height=4, echo=FALSE}
# Creating plot with colour representing each population group
ggplot(monthly_shelter_usage, mapping = aes(x = date, y = actively_homeless, colour = population_group)) +
geom_line() +
labs(x = "Date", 
     y = "Actively Shelter Usage", 
     title =  "Chronic shelter usage remains stable through COVID-19",
     caption = "Began tracking Indigenous numbers in Jan 2021",) +
theme_minimal()
```

@@

Low female representation in homelessness numbers could be due to a variety of factors. Women and children are often especially concerned about the possibility of conflict and abuse due to the public nature of homeless shelters. This would be of even greater concern for victims of sexual abuse that are seeking refuge or are temporarily homeless, may intentionally avoid homelessness shelters or services [@citepallet]. This is an extremely expansive topic that could be an area of further exploration as discussed in Section \@ref(further).

@@

Figure \@ref(fig:graph3) shows the number of people moving to and from permanent housing during the two year period. The red line indicates the number of people leaving the shelter system by obtaining permanent housing and the blue line represents the number of people reentering the shelter system after moving to permanent housing.

```{r graph3, fig.cap="Movement between Permanent housing", echo=FALSE, fig.height=4, fig.width=6}
# Creating plot with two lines, one for moved and one for returned
ggplot(total_population, aes(x = date)) +
geom_line(aes(y = moved_to_housing, col = "Moved")) +
geom_line(aes(y = returned_from_housing, col = "Returned")) + 
labs(x = "Date", 
     y = "# of people", 
     title =  "People moving to permanent housing dropping significantly",
     col = "Action: ") +
theme_minimal() +
theme(legend.position = "bottom") # Putting legend on the bottom for aesthetic reasons 
```
